A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)

### 2014-02-19 text1.tex
proofread, with some open questions

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\geometry{a4paper}                   		% ... or a4paper or a5paper or ... 
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage{amssymb}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}% http://ctan.org/pkg/multirow
\usepackage{tablefootnote}
\usepackage{import}
\usepackage{subcaption}
\usepackage{caption}
\usepackage[format=hang,font=small,labelfont=bf]{caption}

\title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)}
\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth}
%\date{}							% Activate to display a given date or no date

@@

\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field. With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging. 
Hadoop, emerged in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes. We here report a quantitative comparison of Hadoop with regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,Gigabases. In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts. Observing the scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100 Gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution. As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in future biological data analysis.
}

\section{Introduction}
\label{sectionI}
%The problem: dealing with BIG data
Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}. Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.  As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.  Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.

%Current practices: HPC systems
A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample. Most of the bioinformatic programs are written for the Linux operating system. Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made up with distributed computing in mind; many others do not even have the native ability to use multiple cores on the same computer\footnote{Samtools  acquired this feature since v.0.1.19}.

@@

The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node which usually does not exceed~16.  For the tools which do not support the OMP natively, i.e. Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating to split  the dataset and  use multi-node parallelisation.

Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since all the threads synchronisation (or load balancing) has to be coded by a programmer and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.

Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting e.g. involve already existing utilities and cluster tools, split the data into chunks and process them on the separate nodes, and merging the results afterwards. This kind of solution benefits from both MPI-like  and OMP and provides good performance, but the development requires substantial expertise in order to be efficient, and the process is tightly coupled to the local computational cluster and network architectures and in many cases cannot be re-used in other settings.

%Introduce Hadoop, MR, and HDFS
%Properties of Hadoop and HDFS- data localization and programming model etc
The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running task in a {\it massively} parallel way. This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
The prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. 
The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}. Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively moving the computation to the data and reducing network traffic. HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.

%Manuscript focus: When is Hadoop currently an appealing alternative
%Introduce what we did in the manuscript.
In this manuscript we focus on the question {\it if} and {\it when} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective. We use five datasets with short reads of different sizes, align them against the reference genome, and call the variances.
For the existing Hadoop  software we propose modifications, in order to benefit fully from massively parallel nature of MR computations. 
For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short reads alignment and exploiting the network fully, thereby speeding up calculations.
The execution times for each dataset were collected and scaling relations were analysed to answer the question in focus.

The manuscript is structured as follows: In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used. Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources. The results are discussed in Section~IV, and conclusions in Section~V. Supplementary materials are provided in the corresponding section.

@@

\subsection{Datasets}
We used publicly available DNA-seq datasets~(I--III) and one synthetic dataset~(IV) for {\it A.thaliana}, the well known model plant, and one, (dataset~V) for two {\it H.sapiens} individuals (Table~\ref{table:datasets}). Data for datasets I-III and V were obtained using Illumina/(HiSeq) sequencing platforms. Further information about the datasets is provided in the Supplementary material section.

@@

\subsection{Analysis pipelines}
We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and one on regular HPC with a batch processing system (hereafter referred to as HPC). Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage. All experiments were repeated several times and the results averaged to obtain a data point for the particular combination of data size and computational platform. Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (i.e. GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass the same stages on Hadoop and HPC:

@@

In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools. The Bowtie aligner natively implements~OMP meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core. Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions. Where available these features were used to improve the analysis speed. For the exact workflow used one can refer to the code repository created for this work~\cite{code_repo}.

The equivalent Hadoop-based pipeline was implemented with Crossbow. The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments. Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way. Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation.

\subsection{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX, having nodes equipped with 8 dual-core CPUs. Data and reference genomes were stored on a parallel shared storage system\cite{gulo}. The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was set up with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}. For details on computational resources, see Supplementary material.

@@

\subsection{Modified preprocessing stage}
A native preprocessing stage in the Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster.  Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}. The only way to introduce parallelisations in the native preprocessing stage~-- is to split the read files into smaller chunks and to generate a manifest file listing all these chunks. This, however, is lacking the massively parallel way of treating the data.

Being  pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both~--  storage and  computers,  
therefore considered the data already been downloaded to the storage Hadoop can access.
We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
The demand is that the FASTQ data are BZIP2 archived, and  accessible by Hadoop. For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested to the HDFS.
The BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}. Also the BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e. for the developer there is no difference in dealing with the FASTQ data or with its BZIP2 archive.

The Hadoop streaming library offers a possibility to write  Mapper and Reducer for Hadoop jobs in any  programming language. Our Python scripts efficiently process  short reads, produced by  Illumina sequencing platform of different versions, as well as the FASTQ files converted from the SRA format, NCBI Sequence Read Archive. 
The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
To ensure that the both forward and reverse read of the same pair will end up on the same Reducer, the secondary Mapper key-sorting mechanisms were involved.

In order to compare the native preprocessor and the proposed one, the data was put on the HDFS in beforehand, to eliminate possible network delays i.e. while downloading data from some Amazon cloud.
Table~\ref{table:preprocess} illustrates the benefits of our approach.
\begin{table}[htdp]
\small
\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on~$p=56$ cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node, $p=16$ cores.}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
\hline
Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1$	\\
Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
BASH, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
\end{tabular}
\end{center}
\label{table:preprocess}
\normalsize
\end{table}%
To complete the comparison the same preprocessing was performed with the  BASH scripting against the data located on the HPC storage. The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes  multiple cores, with the limiting factor being the  HDD IO performance, which effectively leaves with 5-7 cores of 16.

\subsection{Accuracy of pipelines}
Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp correspondingly) we should not expect them to deliver perfectly matching SNPs lists, but still we expect them to capture and correctly identify the mutation. We tested the correctness just for the smallest dataset~I. The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.

\subsection{Scalability of HPC and Hadoop approaches}
To demonstrate the scalability of  the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}).
As one can see the aligning process with the Bowtie scales fairly good, but the SNP calling part is a bottleneck,  scaling worse than the Bowtie, and consuming progressively larger part of the total calculation time
%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance. During its sorting phase the lack of RAM results in spilling to the HDD. For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our tasks we used the default Samtool's settings.}.  
For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for  the HPC and Hadoop approaches were collected (Table~\ref{table:4}).

One of the attractive sides of using Hadoop~MR is its almost linear scalability,~i.e. calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie etc.). 
Figure~\ref{fig:fig1}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ cores Hadoop cluster. Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}.
To stress the linear nature of scaling, both sets of points were fit to linear polynomial using the least squares method.

\begin{table}[htb]
\small
\caption{Calculation time for Dataset I executed on different  number of cores on a node of  HPC cluster. }
\begin{center}
\begin{tabular}{|c|c|c|c|ccc|}
$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c|}{speed-ups} \\
\hline
	& mapping 	&	SNP calling\tablefootnote{We  used here special tricks how to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781/
}}	&	total  &total & map& SNP call\\
\hline
1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
\end{tabular}
\end{center}
\label{table:3}
\normalsize
\end{table}

@@

\subsection{Used datasets}

>>>>

\documentclass[11pt, oneside]{article}   	% use "amsart" instead of "article" for AMSLaTeX format
\usepackage{geometry}                		% See geometry.pdf to learn the layout options. There are lots.
%\geometry{letterpaper}                   		% ... or a4paper or a5paper or ... 
\geometry{a4paper}                   		% ... or a4paper or a5paper or ... 
%\usepackage[parfill]{parskip}    		% Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}				% Use pdf, png, jpg, or epsБ•√ with pdflatex; use eps in DVI mode
								% TeX will automatically convert eps --> pdf in pdflatex		
\usepackage[normalem]{ulem}
\usepackage{amssymb}
\usepackage[colorlinks]{hyperref}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{multirow}% http://ctan.org/pkg/multirow
\usepackage{tablefootnote}
\usepackage{import}
\usepackage{subcaption}
\usepackage{caption}
\usepackage[format=hang,font=small,labelfont=bf]{caption}

\newcommand{\COMMENT}[1]{{\color{red} #1 }}

\title{A quantitative assessment of the Hadoop framework for analysing massively parallel DNA sequencing data (Hadoop Or Not Hadoop)}
\author{Alexey Siretskiy,  Luca Pireddu, Ola Spjuth \COMMENT{[specify affiliation]}} 
%\date{}							% Activate to display a given date or no date

@@

\abstract{New high-throughput technologies such as massively parallel sequencing has transformed the life sciences into a data-intensive field. With increasing data volumes comes the necessity to analyse data in parallel using high-performance computing resources, but doing this effectively can be laborious and challenging. 
Hadoop, emerging in the last decade, is a framework that automatically distributes data and computation and has been shown to scale to thousands of nodes. Herein we report a quantitative comparison of Hadoop to regular high-performance computing resources for aligning short reads and calling variants for five datasets of different sizes up to 250\,gigabases. In order to increase performance of existing software and obtain a better comparison we modified and wrote new analysis scripts. From the observered scaling relations we are able to draw conclusions about the perspectives of the approaches, leading to the conclusion that as data set sizes reach 100\,gigabases, the Hadoop-based pipelines become performance-competitive with a canonical high-performance cluster solution. As data sets in biological sequencing are sure to increase with time, Hadoop and similar frameworks are very interesting technologies that we envision will play a key role in the future of biological data analysis. \COMMENT{Brit spelling assumed since that's most common in this document}
}

\section{Introduction}
\label{sectionI}
%The problem: dealing with BIG data
Since its inception, massively parallel DNA sequencing, also referred to as Next Generation Sequencing~(NGS) technology, has been an extremely bountiful source of data giving insight into the workings of biological machinery~\cite{metzker, marx}. Decreasing sequencing costs facilitates and promotes larger and larger studies with increasingly larger data sizes, and extracting useful information from these voluminous amounts of data is transforming biology into a data-intensive discipline.  As an example of the scale of the demands, consider that a single Illumina high-throughput sequencing run produces approximately 3\,TB of raw data in 10 days~\cite{illumina}.  Indeed, the Swedish UPPMAX\footnote{\url{uppmax.uu.se}} \COMMENT{[unusual to use both footnotes and endnotes]} high-performance computing (HPC) center recently disclosed data showing that just in their sequencing context (most sequencing performed in Sweden) \COMMENT{huh?} storage is being occupied at a rate of~1\,TB/day while the analyses are using over~1 million computing core-hours per month~\cite{lampa}.

%Current practices: HPC systems
A common step of NGS data analysis consists of mapping short reads to a reference sequence and then finding the genetic variations specific to the sample. \sout{Most of the bioinformatic programs are written for the Linux operating system.} /COMMENT{awkward but maybe it's needed} Some of the most widespread tools\footnote{average number of invocations per month during 2013 at UPPMAX: Samtools~-- 30000, BWA~-- 25000, Bowtie~-- 5000; from internal communications} like BWA~\cite{bwa}, Bowtie~\cite{bowtie} and Samtools~\cite{samtools} are ``regular'' computer programs, not made with distributed computing in mind. Many others do not even have the native ability to use multiple cores \sout{on the same computer}\footnote{Samtools acquired this feature in v.0.1.19}.

@@

The most common approach to speed up NGS tools is to parallelise within a compute node using shared memory parallelism~(OMP)~\cite{openmp}, but this approach is naturally limited by the number of cores per node, which does not usually exceed~16.  For the tools which do not support the OMP natively, e.g. Samtools, variant calling can be parallelised by creating a separate process  for  each chromosome, or using GNU Parallel~\cite{gnuparallel} Linux utility.
Of great importance is that a multi-core approach does not improve the performance of operations that are limited by disk or network throughputs, motivating \COMMENT{huh?} to split the dataset and use multi-node parallelisation.

Message Passing Interface~(MPI)~\cite{mpi1} is a common way to implement multi-node parallelisation, but writing efficient MPI-programs for hundreds of cores is a non-trivial task since thread synchronisation (or load balancing) has to be woven into the software code \sout{by a programmer} and there are only a few existing solutions available for processing sequencing data~\cite{pmap, erne, gnumap}.

Another common way to introduce parallelisation to NGS analysis pipelines in Linux systems is to use Bash scripting. This involves using existing utilities and cluster tools to split the data into chunks, process them on the separate nodes, and merge the results afterwards. This kind of solution benefits from both MPI-like and OMP parallelisation and provides good performance, but the development requires substantial expertise in order to be efficient. Since the process is tightly coupled to the local computational cluster and network architectures, it might not be possible to be re-used in other settings.

%Introduce Hadoop, MR, and HDFS
%Properties of Hadoop and HDFS- data localization and programming model etc
The Map-Reduce~(MR) programming paradigm~\cite{hadoop} offers a compelling alternative for running tasks in a {\it massively} parallel way. This paradigm, however, shifts the focus from the best performance to scalability, suited for managing huge datasets of sizes up to several terabytes~\cite{lin2010}.
The most prevalent open source implementation of Map-Reduce is Hadoop~\cite{hadoop,Hadoop:Guide}. \COMMENT{Fix bib error (can't have both editor and author for this ref type)} 
The Hadoop~MR framework provides automatic distribution of computations over many nodes as well as automatic failure recovery (failure of individual jobs or computing nodes by storing multiple copies on different nodes), and automated collection of results~\cite{Hadoop:Guide}. Hadoop Distributed File System~(HDFS) is a complementary component that stores data by automatically distributing it over the entire cluster, writing data blocks onto the local disk of each node and therefore effectively moving the computation to the data and reducing network traffic. HDFS provides a storage system whose bandwidth and size scales with the number of nodes in the cluster~\cite{Sammer:2012}, which is very different from the properties of the usual HPC cluster network architecture.

%Manuscript focus: When is Hadoop currently an appealing alternative
%Introduce what we did in the manuscript.
In this manuscript we focus on the question {\it if} and {\it when} \COMMENT{that's two questions} Hadoop is an appealing alternative to the program tools generally found in~HPC centers for DNA-seq analysis. Since Hadoop is written in Java, which is slower than the standard~HPC programming languages like C or Fortran, we seek to estimate an average data size when it starts to be worthwhile to use Hadoop from a performance perspective. We use five datasets with short reads of different sizes, align them against the appropriate reference genome, and call the variances.
For the existing Hadoop software we propose modifications, in order to benefit fully from massively parallel nature of MR computations. 
For the classical HPC DNA-seq analysis programs we developed a set of Bash scripts utilizing multiple nodes for short read alignment and exploiting the network fully, thereby speeding up calculations.
The execution times for each dataset were collected and scaling relations were analysed to answer the question\COMMENT{s} in focus.

The manuscript is structured as follows: \COMMENT{[odd to do this in a normal publication, but OK]} In Section~II we briefly introduce the datasets, computational facilities, analysis pipelines design,  and the software used. Then, Section~III presents experimental results; first verifying that both HPC and Hadoop approaches extract the same mutation and then investigating the scaling relations in terms of data size and computing resources. The results are discussed in Section~IV, and conclusions in Section~V. Supplementary materials are provided in the corresponding section.

@@

\subsection{Datasets}
We used publicly available DNA-seq datasets~(I--III), a synthetic dataset~(IV) of {\it A.thaliana}, the well-known model plant, and dataset (V) of two {\it H.sapiens} individuals (Table~\ref{table:datasets}). Data for datasets I-III and V were obtained using Illumina/HiSeq sequencing platforms. Further information about the datasets is provided in the Supplementary material section.

@@

\subsection{Analysis pipelines}
We constructed two pipelines for identifying single-nucleotide polymorphisms (SNPs) from short read data, one based on Hadoop and the other on regular HPC with a batch processing system (hereafter referred to as HPC). Our experiments then consisted of running the pipelines for the selected input dataset and measuring the wall-clock run time for each pipeline stage. All experiments were repeated several times and the results averaged \COMMENT{considering the variations in table 4, why not use medians} to obtain a data point for the particular combination of data size and computational platform. Acknowledging that there are different approaches and software for conducting bioinformatic analysis for HPC (e.g., GATK~\cite{gatk}), we decided to create the  analysis pipeline as simple as possible to be able to pass \COMMENT{huh?} the same stages on Hadoop and HPC:

@@

In the HPC pipeline, reads were aligned with Bowtie, followed by sorting the mapped reads and~SNP calling with Samtools. The Bowtie aligner natively implements~OMP, meaning that with~8 cores on the same computer the result can be theoretically obtained 8 times faster than on a single core. Likewise, Samtools (as of version~0.1.19) also offers shared memory parallelism for several of its functions. Where available these features were used to improve the analysis speed. The exact workflow used is available in the code repository created for this work~\cite{code_repo}. \COMMENT{missing ref}

The equivalent Hadoop-based pipeline was implemented with Crossbow. The input data and the indexed genome reference were copied to Hadoop's storage system~(HDFS) before starting the experiments. Crossbow implements a short pipeline that pre-processes the input data, transforming it into a format suitable for the alignment stage, and then continues to use Bowtie for alignment and SoapSNP to call SNPs.  Unfortunately, Crossbow's preprocessor is not written in MR manner, and thus cannot be run in a massively parallel way. Due to this limitation, this basic step threatened to be the most time-consuming procedure in our test pipeline and bias our experiments. To overcome this bottleneck we substituted Crossbow's preprocessor with our own MR implementation.

\subsection{Computational resources}
To run the HPC analysis pipeline we used a computational cluster at UPPMAX on nodes equipped with 8 dual-core CPUs. Data and reference genomes were stored on a parallel, shared-storage system\cite{gulo}. The Hadoop test platform was deployed on a private cloud at UPPMAX using the OpenNebula~\cite{opennebula} cloud manager. The cluster was set up with Cloudera Hadoop distribution version~2.0.0-mr1-cdh4.5.0~\cite{cloudera}. For details on computational resources, see Supplementary material.

@@

\subsection{Modified preprocessing stage}
A native preprocessing stage in Crossbow\footnote{the same holds for Myrna, a cloud-based solution for differential expression analysis} provides great flexibility in delivering the data to the Hadoop cluster. Data can be downloaded from Amazon S3, FTP and HTTP servers over the Internet\cite{crossbow}. The only way to introduce parallelisations in the native preprocessing stage is to split the read files into smaller chunks and to generate a manifest file listing all these chunks. This, however, is lacking the massively parallel way of treating the data.

Being pragmatic we assume that sequencing platforms are usually affiliated with computation facilities which provide both storage and computers, therefore we considered that the data already been downloaded to storage Hadoop could access.
We rewrote the preprocessing stage scripts in a MR-manner, benefiting from its massively parallel nature.
The requirement is that the FASTQ data are BZIP2 archived and accessible by Hadoop. For our case the storage with the data was mounted with SSHFS to one of the Hadoop nodes, from where the BZIP2'ed data were ingested \COMMENT{special word?} to the HDFS.
BZIP2 provides very good compression and is splittable, meaning that the archive can be expanded in a parallel manner\footnote{other options like splittable LZO are also possible}. Also, BZIP2 format is natively supported by Hadoop by enabling the respective codec, i.e., \sout{for the developer} there is no difference in dealing with the FASTQ data or with its BZIP2 archive.

The Hadoop streaming library offers a possibility to write Mapper and Reducer for Hadoop jobs in any programming language. Our Python scripts efficiently process short reads, produced by Illumina sequencing platform of different versions, as well as the FASTQ files converted from SRA format (NCBI Sequence Read Archive).\COMMENT{ref?} 
The problem of lacking the unique FASTQ header standard was solved by the rewriting the header based on a SHA-1 hash function, and the reads mates were labeled with ``.1" and ``.2" suffixes. 
To ensure that both the forward and reverse reads of the same pair would end up on the same Reducer, secondary Mapper key-sorting mechanisms were involved.

In order to compare the native preprocessor and the proposed one, the data was put on the HDFS beforehand, to eliminate possible network delays, e.g., while downloading data from some Amazon cloud.
Table~\ref{table:preprocess} illustrates the benefits of our approach.
\begin{table}[htdp]
\small
\caption{Timings (in minutes) with the  Crossbow native read preprocessor and with derived approach for  datasets II, III, IV  on \sout{$p=56$} 56 cores Hadoop cluster, using the Crossbow, and with the BASH script on a single HPC node \sout{, $p=16$} with 16 cores. \COMMENT{A bit odd to use the vertical lines before/after table, but OK.}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|}
Dataset			&		II (7 Gbases )	& III (30 Gbases)	& IV (100 Gbases)\\
\hline
Crossbow native			&		$60.6\pm0.7$	& $299.0\pm2.3$	&	$673\pm1.0$	\\
Hadoop, this work			&		$7.0\pm0.0$	&	$20.7\pm0.4$&		$52.4\pm0.1$\\
BASH, this work			& 		$7.4\pm0.0$	&	$31.1\pm0.1$	&	$114.5\pm0.3$	\\
\end{tabular}
\end{center}
\label{table:preprocess}
\normalsize
\end{table}%
To complete the comparison the same preprocessing was performed with the BASH scripting against the data located on the HPC storage. The script\footnote{\texttt {pbzip2 -dc  \${file1} | paste - - - - -d'\textbackslash t' | cut -f1,2,4 | paste - -d' ' <(pbzip2 -dc  \${file2} | paste - - - - -d'\textbackslash t' | cut -f2,4) | pbzip2 -cz > \${fileOut}}} ran on a single HPC node, utilizes multiple cores, with the limiting factor being the HDD IO performance, which effectively leaves with 5-7 cores of 16. \COMMENT{awkward}

\subsection{Accuracy of pipelines}
Since our HPC and Hadoop approaches use different~SNP callers (Samtools and SOAPsnp, correspondingly) we should not expect them to deliver perfectly matching SNP lists, but still we expect them to capture and correctly identify the mutations. We tested the correctness just for the smallest dataset (I). The mutation $C\rightarrow T$ on chromosome 4 at position $16702262$\cite{schneeberger} was successfully localized by both applications.

\subsection{Scalability of HPC and Hadoop approaches}
To demonstrate the scalability of the HPC approach we collected running times for dataset~I as a function of the number of cores used (Table~\ref{table:3}). \COMMENT{Consider replacing with a graph.}
\sout{As one can see} the aligning process with the Bowtie scales fairly well, but the SNP calling part is a bottleneck, scaling worse than the Bowtie, and consuming a progressively larger portion of the total calculation time.
%\footnote{The amount of available memory, specified with the \texttt{-m} option, is of big importance for the Samtools performance. During its sorting phase the lack of RAM results in spilling to the HDD. For example for an half of  Dataset III, SNP calling takes approximately 40 minutes with 8 GB RAM, and 15 minutes with 48 GB RAM, when  the whole BAM file fits into RAM. For our tasks we used the default Samtool's settings.}.  
For the given datasets (I--V) the timings for alignment against the corresponding genomes and SNP calling for the HPC and Hadoop approaches were collected (Table~\ref{table:4}).

One of the attractive sides of using Hadoop~MR is its almost linear scalability, i.e., calculation time linearly depends on the size of the dataset\cite{crossbow,seal}, regardless the scaling nature of the underlying program (Samtools, Bowtie, etc.).
Figure~\ref{fig:fig1} \COMMENT{this figure should be closer, table and figure layout odd in general}, based on Table~\ref{table:4}, shows the calculation time as a function of the dataset size for $p=56$ \COMMENT{is this 'p=' notation necessary} cores Hadoop cluster. Dataset V was excluded since it is for {\it H.sapiens}, which has more than 20 times larger genome than~{\it A.thaliana}. \COMMENT{so what}
To stress the linear nature of scaling, both sets of points were fit \sout{to linear polynomial} using the least squares method.

\begin{table}[htb]
\small
\caption{Calculation time for Dataset I executed on different number of cores on a node of HPC cluster. \COMMENT{awkward title, consider replacing with graph (and new title)}}
\begin{center}
\begin{tabular}{|c|c|c|c|ccc|}
$N$ cores	&\multicolumn{3}{c|}{timing, minutes}&\multicolumn{3}{c|}{speed-ups} \\
\hline
	& mapping 	&	SNP calling\tablefootnote{We used special tricks to parallelize the Samtools analysis by chromosome, as exemplified here \url{http://www.biostars.org/p/48781}.}	&	total  &total & map& SNP call\\
\hline
1	&	22.75	$\pm$1.24&	26.1$\pm$0.7	&	48.9	&	 1.00	&	1.00	&	1.00\\
2	&	11.00$\pm$0.05	&	16.9$\pm$1.3	&	27.9	&	 0.88	&	1.03	&	0.77\\
4	&	5.91$\pm$0.04	&	11.7$\pm$0.5	&	17.6	&	 0.69	&	0.96	&	0.56\\
6	&	4.24$\pm$0.04	&	8.6$\pm$0.8	&	12.8	&	 0.64	&	0.89	&	0.51\\
8	&	3.44$\pm$0.03	&	8.6$\pm$0.9	&	12.1	&	 0.51	&	0.83	&	0.38\\
10	&	2.91$\pm$0.01	&	7.7$\pm$0.5	&	10.6	&	 0.46	&	0.78	&	0.34\\
12	&	2.64$\pm$0.04	&	7.5$\pm$0.5	&	10.1	&	 0.40	&	0.72	&	0.29\\
14	&	2.57$\pm$0.03	&	7.5$\pm$0.6	&	10.1	&	 0.35	&	0.63	&	0.25\\
16	&	2.88$\pm$0.06	&	7.4$\pm$0.4	&	10.2	&	 0.30	&	0.49	&	0.22\\
\end{tabular}
\end{center}
\label{table:3}
\normalsize
\end{table}

@@

\subsection{Datasets}
